Broom: Safe Region-Based Memory Management for Dataflow Programs

### 2015-11-11 paper/type-inference.tex
Edits.

% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
In general, validity constraints generated by our algorithm are of the
form:
\begin{center}
\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {\phicstr \conj \bigwedge_{j}F_j(\varphi_j)}\)
\end{center}
Where $\varphi_i$'s denote the unknown preconditions of the class or
the method under elaboration. If the constraint is generated while
checking the well-formedness of a type or elaborating an expression,
then $\varphi_j$'s denote the unknown preconditions of classes and
methods that were used in that type or expression.  Each use of a
(region-polymorphic) class or a method may instantiate region
parameters differently, resulting in a different pending substitution
($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
of predicate variables and pending substitutions.  They capture the
concrete relationships between regions that are \emph{known} to hold,
and \emph{required} to hold, respectively, under the context in which
the constraint was generated.

@@

Next, we convert the dependency graph over constraints into a
dependency DAG ($G_C$) over sets of constraints, where each set
represents a strongly connected component in the dependency graph.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph. Accordingly, we
obtain a topological ordering of nodes in the graph $G_{{C}}$, and
solve the sets of constraints in that order. The solutions obtained
after solving a constraint set are applied to the constraints in
subsequent sets before attempting to solve them. Consequently, when
the turn of a constraint set ($C$) arrives during the constraint solving
process, it satisfies certain properties:
\begin{itemize}
\item There exists only one predicate variable that is either
constrained or used by the constraints in the set ($C$). The variable is
called set's \emph{subject}. This property follows from (a). the fact
that all the dependency constraints have already been solved (and
solutions applied), and (b). the assumption that there are no mutually
recursive definitions. 
\item All the constraints that constrain the set's subject are present
in the set. This follows from our definition of the dependency relation.
\end{itemize}

Solving the set ($C$) of constraints entails finding an assignment for
$C$'s subject, and also all the region variables that occur free in
$C$, such that the solutions satisfy the well-formedness constraints.
We now formalize the constraint satisfaction problem, and its
solution.

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem})}
A constraint satisfaction problem is a tuple
$(C,\rhoenv,\varphi,\bar{\rho} \in \bar{\rhoenv})$, where $C$, the set
of validity constraints, is of the form:
\begin{center}
\(
  \bigcup_{i=1}^n \{
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i} \conj F_i(\varphi)}
  \}
\)
\end{center}
The subject of the set is $\varphi$, and its free region variables
range over $\rhobar$. The solution to the constraint satisfaction
problem is a pair $(\substFn,\phisol)$, where $\substFn$ is a map from
region variables to region identifiers such that $\substFn(\rho_j) \in
\rhoenv_j$, for every $j$, and $\phisol$ is a constraint formula such
that:
\begin{itemize}
\item $\phisol$ is well-formed under $\rhoenv$ (i.e.,
$\tywf{\rhoenv}{\phisol}$).
\item $\phisol$ satisfies every validity constraint in $\substFn(C)$.
That is, the following sequent is valid for every $i$:
\begin{center}
\(
  \isvalid{\substFn(\phictxt^{i}) \conj \phisol}
          {\substFn(\phicstr^{i}) \conj (\substFn \circ F_i)(\phisol))}
\)
\end{center}
\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
that $\phisol'$ satisfies every validity constraint in $\substFn(C)$,
and $\phisol'$ is strictly weaker than $\phisol$ (i.e.,
$\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
\end{itemize}
\end{definition}

\subsection{The Constraint Satisfaction Problem}

We now formally define the constraint satisfaction problem over sets
of constraints. For simplicity, we ignore mutual recursion and assume
that a a set only contains (possibly recursive) constraints over a
single predicate variable. The general form of such a set of
constraints is as following:
\begin{center}
\(
  \bigcup_{i=1}^n \{
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i} \conj \bigwedge_{k=1}^n F_k(\varphi)}
  \}
\)
\end{center}
% The solution to the above set of constraints is a tuple $(\substFn,)$
% problem is to find a map

% After solving all the dependencies and applying solutions, the most
% general form of a set of constraints is the following:
% \begin{center}
% \(
%   \bigcup_{i=1}^n \{
%     \bigcup_{j=1}^m \{ 
%       \isvalid{\phictxt^{ij} \conj \varphi_i}{\phicstr^{ij} \conj
%       \bigwedge_{k=1}^n F_k(\varphi_k)}
%       \}
%     \}
% \)
% \end{center}
% Our technical report~\cite{techrep} contains formal proofs
% establishing the equivalence of the above constraint set and the one
% below, with respect to the constraint satisfaction problem:
% \begin{center}
% \(
%   \bigcup_{i=1}^n \{
%     \isvalid{\bigwedge_{j=1}^m \phictxt^{ij} \conj \varphi_i}
%             {\bigwedge_{j=1}^m \phicstr^{ij} \conj \bigwedge_{k=1}^n F_k(\varphi_k)}
%   \}
% \)
% \end{center}
% Our technical report~\cite{techrep} contains formal proofs

\subsubsection{Solving Non-Recursive Constraints}

Non-recursive constraints are of the form $\isvalid{\phictxt \wedge
\varphi}{\phicstr}$, where the predicate variable $\varphi$ has an
associated well-formedness constraint $\tywf{\rhoenv}{\varphi}$.  The
formula $\phicstr$ may contain free region variables ($\rho$), each
with its own well-formedness constraint ($\rho \in \rhoenv$)
specifying its domain of unification. Solving the constraint entails
generating well-formed assignments to $\varphi$ and all free region
variables in $\phicstr$.

\begin{definition}
\emph{(\textbf{\csolve})}
\label{def:csolve}
Let $L(S,R_1,R_2)$ be the following constraint language:
\begin{smathpar}
\begin{array}{lcl}
x,y & \in & S\\
\phi & \coloneqq & true \ALT R_1(x,y) \ALT R_2(x,y) \ALT \phi \conj \phi \\
\end{array}
\end{smathpar}
\csolve is a function that accepts:
\begin{itemize}
\item A set $C$ of constants, and a set $V$ of variables
\item A constraint $\phi$
\end{itemize}

\end{definition}

>>>>

% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
In general, validity constraints generated by our algorithm assume one
of the following two forms:
\begin{center}
\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {\phicstr} \qquad
   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {F_j(\varphi_j)}
\)
\end{center}
Where $\varphi_i$'s denote the unknown preconditions of the class and
the method under elaboration. If the constraint is generated while
checking the well-formedness of a type or elaborating an expression,
then $\varphi_j$'s denote the unknown preconditions of classes and
methods that were used in that type or expression.  Each use of a
(region-polymorphic) class or a method may instantiate region
parameters differently, resulting in a different pending substitution
($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
of predicate variables and pending substitutions. While $\phictxt$
captures relationships that are \emph{known} to hold between concrete
region identifiers (i.e., $\rgn$'s) when the constraint was generated,
$\phicstr$ captures relationships that are \emph{required} to hold
among region varibles (i.e., $\rho$'s), or relationships between
region variables and identifiers. Each region variable occuring in
$\phicstr$ has an associated well-formedness constraint, which
specifies its unification domain. The unification domain of a
constraint is the union of unification domains of all region variables
occuring in the constraint.

@@

Next, we convert the dependency graph over constraints into a
dependency DAG ($G_C$) over sets of constraints, where each set
represents a strongly connected component in the dependency graph.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph. Accordingly, we
obtain a topological ordering of nodes in the graph $G_{{C}}$, and
solve the sets of constraints in that order. The solutions obtained
after solving a constraint set are applied to the constraints in
subsequent sets before attempting to solve them. Consequently, when
the turn of a constraint set ($C$) arrives during the constraint solving
process, it satisfies certain properties:
\begin{itemize}
\item There exists only one predicate variable ($\varphi$) that is either
constrained or used by the constraints in the set ($C$). The variable is
called set's \emph{subject}. This property follows from (a). the fact
that all the dependency constraints have already been solved (and
solutions applied), and (b). the assumption that there are no mutually
recursive definitions. 
\item All the constraints that constrain the set's subject are present
in the set. This follows from our definition of the dependency relation.
\end{itemize}

Solving the set ($C$) of constraints entails finding an assignment for
$C$'s subject ($\varphi$), and also all the region variables
($\rhobar$) that occur free in $C$, such that the solution satisfies
well-formedness constraints on $\varphi$ and $\rhobar$. To simplify
presentation, we think of $C$ as being parameterized on
$\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
\rbrack$. We now formalize the constraint satisfaction problem, and
its solution.

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem (CSP)})} A constraint
satisfaction problem is a tuple $\CSP$, where $C\lbrack
\varphi,\rhobar \rbrack$ is a set of validity constraints, where
$i$'th validity constraint assumes one of the following forms:
\begin{center}
\(
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i}}\qquad
    \isvalid{\phictxt^{i} \conj \varphi}
            {F_i(\varphi)}
\)
\end{center}
$\bar{\rhoenv}$ are the unification domains for $\rhobar$. We call the
union of all unification domains ($\bigcup\bar{\rhoenv}$) as the
unification domain ($\rhoenv$) of the CSP.  The solution to the
constraint satisfaction problem is a pair $(\substFn,\phisol)$, where
$\substFn$ is a map from $\rhobar$ to $\rhoenv$ such that
$\substFn(\rho_j) \in \rhoenv_j$, for every $j$, and $\phisol$ is a
constraint formula such that:
\begin{itemize}
\item $\phisol$ is well-formed under $\rhoenv_{\varphi}$ (i.e.,
$\tywf{\rhoenv_{\varphi}}{\phisol}$).
\item Every sequent in $C\lbrack (\phisol,\substFn(\rhobar)) \rbrack$
is valid.
% \begin{center}
% \(
%   \isvalid{\phictxt^{i} \conj \bigwedge_j(\rho_j = \substFn(\rho_j)) \conj \phisol}
%           {\phicstr^{i} \conj  F_i(\phisol))}
% \)
% \end{center}
\item $\phisol$ is maximally weak. That is, $\nexists \phisol'$ such
that $\phisol'$ is well-formed, $C\lbrack (\phisol',\substFn(\rhobar))
\rbrack$ is valid, and $\phisol'$ is strictly weaker than $\phisol$
(i.e., $\isvalid{\cdot}{\phisol \Rightarrow \phisol'}$ but
$\isnotvalid{\cdot}{\phisol' \Rightarrow \phisol}$).
\end{itemize}
\end{definition}

\subsection{Solving the CSP}

The first step of solving the CSP $\CSP$ is to cast it as an
equivalent problem $\csp$ involving a single validity constraint. The
single constraint ($\lbrack \varphi,\rhobar \rbrack$) is:
\begin{center}
\(
  \isvalid{ \Phicx \conj \varphi}
        {\Phics \conj  \bigwedge_i F_i(\varphi)}
\)
\end{center}
Where $\Phicx = \bigwedge_i \phictxt^{i}$ and $\Phics = \bigwedge_i
\phicstr^{i}$. To see why both CSPs are equivalent, consider two
distinct constraints, $c_i\lbrack \varphi,\rhobar_i \rbrack$ and
$c_j\lbrack \varphi,\rhobar_j \rbrack$. Since $i\neq j$, $\rhobar_i
\neq \rhobar_j$. Without the loss of generality, assume that $c_i$ was
generated before $c_j$ by the constraint generation algorithm. Observe
that when our constraint generation algorithm generates a constraint,
all relationships between concrete region identifiers referred by the
consequent of constraint are already present in its antecedent (this
includes the unification domains of region variables referred by the
consequent).  Since no new relationships between existing regions are
added when a new constraints is generated, $\phictxt^j$ either
describes the same relationships that are already described by
$\phictxt^i$, or describes relationships among region identifiers that
are new, and not relevant to $c_i$. Therefore, strengthening the
context of $c_i$ with that of $c_j$ (or vice versa) neither weakens
nor strengthens $c_i$ (resp.  $c_j$).

\subsubsection{Non-Recursive Constraints}
\label{sec:csolve}

We first describe how we solve a non-recursive CSP $\csp$ where
c$\lbrack \varphi,\rhobar \rbrack$ is of the form $\isvalid{\Phicx
\conj \varphi}{\Phics}$.

Our approach is based on the observation that $\FB$ does not admit
null values or uninitialized variables, thus forcing every region
variable to be unified with some concrete region. The constraint
formula $\Phics$ captures all such unification constraints on
$\rhobar$. Since the set of all concrete region identifiers is the
unification domain ($\rhoenv$) of CSP, this means for every $\rho_i$
with a well-formedness constraint as $\rho_i \in \rhoenv_i$, there
exists a $\rgn \in \rhoenv$ such that and $\isvalid{\Phicx \conj
\Phics}{\rho = \rgn}$. However, $\rgn$ may not belong to $\rhoenv_i$,
in which case the well-formedness constraint on $\rho_i$ is not
satisfied, and constraint solving must fail.  Therefore, there exists
a unique assignment $\substFn$ such that $c\lbrack
\phi_{sol},\substFn(\rhobar)$ is satisfied, regardless of
$\phi_{sol}$.

To obtain $\phisol$, we make use of another observation. Let $c\lbrack
\varphi, \substFn(\rhobar) \rbrack$ be the following constraint:
\begin{center}
\(
  \isvalid{ \Phicx \conj \varphi}
        {\Phics'}
\)
\end{center}
Consider a maximally weak formula $\phi$ such that
$\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$. 
% Since $\phi$ is
% maximally weak, there does not exist a $\phi'$ satisfying all the
% below three conditions:
% \begin{itemize}
% \item $\isvalid{\Phicx}{\phi' \Leftrightarrow \Phics'}$.
% \item $\isvalid{\cdot} {\phi \Rightarrow \phi'}$.
% \item $\isnotvalid{\cdot}{\phi' \Rightarrow \phi}$.
% \end{itemize}
Clearly, $\tywf{\rhoenv}{\phi}$. However, since $\rhoenv_{\varphi}
\subseteq \rhoenv$, we have two cases:
\begin{itemize}

\item Case $\tywf{\rhoenv_{\varphi}}{\phi}$: This means that $\phi$ is
a solution to $\varphi$. 
% Furthermore, it is the weakest solution, because it is equivalent to
% $\Phics$ under $\Phicx \conj \bigwedge_j(\rho_j =
% \substFn(\rho_j))$, and any other solution has to stronger than
% $\Phics$ under the same context.

\item Case $\tynwf{\rhoenv_{\varphi}}{\phi}$: In this case, $\phi$
contains at least one equality or outlives constraint on two region
identifiers, $\rgn_i$ and $\rgn_j$, where (a). $\rgn_i,\rgn_j \in
\rhoenv - \rhoenv_{\varphi}$, or (b). $\rgn_i \in \rhoenv -
\rhoenv_{\varphi}$ and $\rgn_j \in \rhoenv_{\varphi}$, such that the
constraint is not implied by $\Phicx$ (if it is implied, then $\phi$
is not maximally weak). Let us denote such constraint on $\rgn_i$ and
$\rgn_j$ as $\phi_{ij}$. Now, let us consider a solution $\phisol$ to
$\varphi$, which means that $\isvalid{\Phicx \conj \phisol}{\Phics'}$.
Since $\isvalid{\Phicx}{\phi \Leftrightarrow \Phics'}$, we have
$\isvalid {\Phicx \conj \phisol}{\phi}$. Since
$\isvalid{\phi}{\phi_{ij}}$, we have $\isvalid {\Phicx \conj
\phisol}{\phi_{ij}}$, although $\isnotvalid{\Phicx}{\phi_{ij}}$. But
this is impossible. To see why, recall that all the constraints on
identifiers in $\rhoenv_{\varphi}$ occuring in $\Phicx$ are subsumed
by $(\rhoenv - \rhoenv_{\varphi}) \outlives \rhoenv_{\varphi}$.
Therefore, it is impossible to derive $\phi_{ij}$ from $\Phicx$ by
only adding constraints on $\rgn \in \rhoenv_{\varphi}$. Hence, such a
solution $\phisol$ cannot exist.
\end{itemize}
The above discussion hints at an algorithm to compute a solution to
$\varphi$: find a maximally weak $\phisol$ such that $\isvalid{\Phicx}
{\phisol \Leftrightarrow \Phics'}$. If $\tywf{\rhoenv_{\varphi}}
{\phisol}$, then $\phisol$ is the solution. Otherwise, there is no
solution to $\varphi$.

\subsubsection{Solving Recursive Constraints}

We now extend constraint solving to recursive CSP $\csp$, where
$c\lbrack \varphi,\rhobar \rbrack$ assumes the form $\isvalid{\Phicx
\conj \varphi}{\Phics \conj \bigwedge_i F_i(\varphi)}$. For the sake
of brevity, we define $G(\varphi) = \Phicx \conj \bigwedge_i
F_i(\varphi)$, and use $F$ in $c\lbrack \varphi,\rhobar \rbrack$: 
\begin{center}
\(
  \isvalid{\Phicx \conj \varphi}{G(\varphi)}
\)
\end{center}
To solve the above recursive constraint, we start with the observation
that the set of all constaints ($\phi$) over $\rhoenv \cup
\{\rhobar\}$ is a lattice, where $\phi_1 \le \phi_2 \triangleq
\isvalid{\cdot}{\phi_1 \Rightarrow \phi_2}$. Note that $G$ is a
monotone over the lattice:
\begin{center}
  $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
  \phi_2) \Rightarrow (G(\phi_1) \Rightarrow G(\phi_2))}$
\end{center}
From Knaster-Tarski's theorem, we know that $G$ has a greatest fixed
point ($\phi_f$), with following properties:
\begin{itemize}
\item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
$\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
\item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
{\cdot} {\phi_f' \Rightarrow \phi_f}$.
\end{itemize}
We therefore compute the greatest fixed point ($\phi_f$) of $G$, and
convert the recursive constraint to the following non-recursive
constraint:
\begin{center}
\(
  \isvalid{\Phicx \conj \varphi}{\phi_f}
\)
\end{center}
The technique described in \S~\ref{sec:csolve} now suffices to solve
the above constraint.

% Note that $F$ is a non-idempotent substitution function with following
% properties:
% \begin{itemize}
% \item $F$ is a monotone: 
% \begin{center}
%   $\forall \phi_1,\phi_2.\; \isvalid{\cdot}{(\phi_1 \Rightarrow
%   \phi_2) \Rightarrow (F(\phi_1) \Rightarrow F(\phi_2))}$
% \end{center}

% \item $F$ distributes over conjunction (mind the syntactic equality):
% \begin{center}
%   $\forall \phi_1,\phi_2.\; F(\phi_1 \conj \phi_2) = F(\phi_1) \conj
%   F(\phi_2)$
% \end{center}
% \end{itemize}

% Define $G(\varphi) = \Phics \conj F(\varphi)$. Note that $G$ is also a
% monotone, hence $G$ has greatest fixed point (GFP) in the lattice. Let
% $\phi_f$ be the GFP of $G$. Following properties ensue:
% \begin{itemize}
% \item \textbf{Property 1} $\phi_f = G(\phi_f)$, which means
% $\isvalid{\cdot}{\phi_f \Leftrightarrow G(\phi_f)}$
% \item \textbf{Property 2} $\forall \phi_f'$ such that $\phi_f' =
% G(\phi_f')$, we have $\phi_f' \le \phi_f$, which means $\isvalid
% {\cdot} {\phi_f' \Rightarrow \phi_f}$.
% \end{itemize}

% \begin{theorem}
% $\phi_f$ is a solution.
% \end{theorem}
% \begin{proof}
% $\phi_f$ is GFP of $G$. Therefore, $\isvalid{\cdot}{\phi_f
% \Leftrightarrow G(\phi_f)}$. By strengthening the context, we get
% $\isvalid{\Phicx}{\phi_f \Leftrightarrow G(\phi_f)}$. It follows that
% $\isvalid{\Phicx \conj \phi_f}{G(\phi_f)}$. Hence, $\phi_f$ is a
% solution for $\varphi$.
% \end{proof}

% \begin{theorem}
% If $\phisol$ is a solution of $\varphi$, then $\isvalid{\Phicx \conj
% \phisol}{\phi_f}$
% \end{theorem}
% \begin{proof}
%   Since $\phisol$ is a solution, $\isvalid{\Phicx}{\phisol \Rightarrow
%   G(\phisol)}$. Therefore, $\isvalid{}{\Phicx \conj \phisol \Rightarrow
%   \Phicx \conj G(\phisol)}$. Therefore, $\isvalid{}{G(\Phicx) \conj
%   G(\phisol) \Rightarrow G(\Phicx) \conj G^2(\phisol)}$

% \end{proof}
